---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
---

# Set-up your environment

```{r Load Packages, echo=FALSE, message=FALSE, warning=FALSE}
library(cbsodataR)
library(dplyr)
library(tidyr)
library(sf)
library(knitr)
library(ggplot2)
```

# Title Page

Include your names
Melvin Hofman
Max Bergsma
Gijs Stevens
Hidde Voors
Imad el Hilali
Jules Bierings


Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Housing affordability has become an increasingly important social issue in the Netherlands over the past decade. House prices have risen substantially in many parts of the country, while income growth have not kept pace. As a result, many households, particularly first-time buyers and younger generations, face greater difficulties entering the housing market, according to DNB (DNB, 2026). In addition, CBS states that housing costs in the Netherlands are higher than most EU countries (CBS, 2024).

Understanding whether house prices have increased faster than incomes is therefore important for evaluating changes in housing affordability. If housing prices grow more rapidly than household incomes, purchasing a home becomes increasingly difficult for a large share of the population. This may have significant social and economic consequences, including increased inequality between homeowners and non-homeowners.

The main research question of this study is:

Have housing prices in the Netherlands increased faster than incomes over the past ten years?

To answer this question, the following sub-questions are examined:
1. How have house prices and average incomes developed between 2014 and 2024?
2. How has housing affordability changed during this period?
3. Are there significant differences in housing affordability between Dutch provinces?
4. Has the introduction of the new rent law changed the relationship between housing prices and income?

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r Download Data, echo=FALSE, message=FALSE, warning=FALSE}
# data_download <- cbs_get_data("70072NED")

# Add region label to the dataset
# raw_dataset <- data_download %>% cbs_add_label_columns()
# write.csv(raw_dataset, "data/raw_dataset.csv")

# Load saved dataset
raw_dataset = read.csv("data/raw_dataset.csv")
```


## 2.2 Provide a short summary of the dataset(s)

The primairy data for this study is sourced from Statistics Netherlands (CBS). The original comprehensive dataset consists of 24,335 observations (rows) and 326 variabeles (columns), capturing key socio-economic indicators such as year, regions, disposable income per employee, WOZ property values.

Given the extensive nature of the data, the single source was programmatically divided into two distinct subsets within R to isolate the specific variabeles required for analysis. The first subset captures detailed data on income per region, while the second focuses on regional WOZ values over time. To reconstruct the data for analysis, these two refined datsets were subsequently merged back together using relational filtering. This process ensures data usability while retaining critical demographic data to evaluate differences across various sub-populations and regions.

```{r Homevalue Dataset, echo=FALSE, message=FALSE, warning=FALSE}
huizenprijzen_provincie = raw_dataset %>%
  mutate(
    Year = as.integer(substr(Perioden, 1, 4))
  ) %>%
  filter(
    Year == 2023,
    startsWith(RegioS, "PV")
  ) %>%
  select(
    Year,
    RegioS,
    RegioS_label,
    GemiddeldeWOZWaardeVanWoningen_98
  )
write.csv(huizenprijzen_provincie, "data/huizenprijzen_provincie.csv")

kable(huizenprijzen_provincie)
```

```{r Income Dataset, echo=FALSE}
inkomen_provincie <- raw_dataset %>%
  mutate(
    Year = as.integer(substr(Perioden, 1, 4))
  ) %>%
  filter(
    Year == 2023,
    startsWith(RegioS, "PV")
  ) %>%
  select(
    Year,
    RegioS,
    RegioS_label,
    BronInkomenAlsWerknemer_141
  )
write.csv(inkomen_provincie, "data/inkomen_provincie.csv")

kable(inkomen_provincie)
```

```{r Province Datasets Merge, echo=FALSE}
provincie_samengevoegd = huizenprijzen_provincie %>%
  inner_join(
    inkomen_provincie,
    by = c("Year", "RegioS", "RegioS_label")
  ) %>%
  rename(
    GemiddeldeWOZ = GemiddeldeWOZWaardeVanWoningen_98,
    InkomenWerknemer = BronInkomenAlsWerknemer_141
  )  %>%
  mutate(statcode = RegioS)

kable(provincie_samengevoegd, digits = 2)
```

```{r Landelijke Dataset, echo=FALSE}
landelijk <- raw_dataset %>%
  mutate(
    Year = as.integer(substr(Perioden, 1, 4))
  ) %>%
  filter(
    Year >= 2015,
    Year <= 2024,
    startsWith(RegioS, "NL")
  ) %>%
  select(
    Year,
    RegioS_label,
    BronInkomenAlsWerknemer_141,
    GemiddeldeWOZWaardeVanWoningen_98
  ) %>%
  rename(
    GemiddeldeWOZ = GemiddeldeWOZWaardeVanWoningen_98,
    InkomenWerknemer = BronInkomenAlsWerknemer_141
  )

write.csv(landelijk, "data/landelijk.csv")

kable(landelijk)
```

```{r Summarize Data, echo=FALSE}
provincie_summary = provincie_samengevoegd %>%
  summarise(
    Mean_WOZ = mean(GemiddeldeWOZ),
    Min_WOZ = min(GemiddeldeWOZ),
    Max_WOZ = max(GemiddeldeWOZ),
    SD_WOZ = sd(GemiddeldeWOZ),
    
    Mean_Income = mean(InkomenWerknemer),
    Min_Income = min(InkomenWerknemer),
    Max_Income = max(InkomenWerknemer),
    SD_Income = sd(InkomenWerknemer),
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Statistic", "Variable"),
    names_sep = "_",
    values_to = "Value"
  ) %>%
  pivot_wider(
    names_from = Statistic,
    values_from = Value
  ) %>%
  mutate(
    Variable = recode(
      Variable,
      WOZ = "Average WOZ Value",
      Income = "Employee Income",
      Affordability = "WOZ-to-Income Ratio"
    )
  ) %>%
  select(Variable, Mean, Min, Max, SD)

kable(
  provincie_summary,
  caption = "Descriptive statistics for home value in provinces in the Netherlands, 2023",
  digits = 2
)
```
## 2.3 Describe the type of variables included

The variabeles included in this study predominantly capture Socio-Economic Status (SES) and housing market characteristics. The key indicators can be broken down into three main categories:
1. SES metrics: Disposable income per employee serves as our primary continuous of household financial capacity.
2. Market metrics: The WOZ property value represents the offical appraised market valuation of residental real estate over time.
3. Demographic metrics: The use of regions and urbanization levels allow for granular sub-group analysis.

Rather than being collected through self-supported individual interviews or surveys, this dataset consist entirely of administrative data collected by Statistics Netherland (CBS). The income and demographic data originate directly from national municipal registries (BRP) and tax authority records, while the housing data is compiled from the national valuation registry (Landelijke Voorziening WOZ).Because this data relies on offical administrative registration rather than sample surveys, it minimizes self-reporting bias and provides comprehensive coverage of the target populations. 


# Part 3 - Quantifying

## 3.1 Data cleaning

To prepare our data for quantitative analysis, we executed several structured cleaning steps. First, provincial location data was originally recorded in numerical codes. We mapped and translated these administrative codes into their respective official Dutch province names. Second, we systematically handled missing values ($NAs$). To safeguard the statistical integrity of our analysis, we selected a time period that is relevant and without any missing values ($NAs$) at other variables.


## 3.2 Generate necessary variables


```{r New Variable 1, echo=FALSE}
provincie_samengevoegd <- provincie_samengevoegd %>%
  mutate(
    AffordabilityRatio = GemiddeldeWOZ / InkomenWerknemer
  )
provincie_samengevoegd

write.csv(provincie_samengevoegd, "data/provincie_samengevoegd.csv")
```
```{r New Variable 2, echo=FALSE}
landelijk = landelijk %>%
  mutate(
    AffordabilityRatio = GemiddeldeWOZ / InkomenWerknemer,
    WOZ_growth_yoy = (GemiddeldeWOZ / lag(GemiddeldeWOZ) - 1) * 100,
    Income_growth_yoy = (InkomenWerknemer / lag(InkomenWerknemer) - 1) * 100,
    GrowthGap_yoy = WOZ_growth_yoy - Income_growth_yoy
  )
kable(landelijk)
```



## 3.3 Visualize temporal variation

```{r Time Plot, echo=FALSE}
# Deze plot gemaakt met chat

landelijk_index <- landelijk %>%
  arrange(Year) %>%
  mutate(
    WOZ_index = GemiddeldeWOZ / GemiddeldeWOZ[Year == 2015] * 100,
    Inkomen_index = InkomenWerknemer / InkomenWerknemer[Year == 2015] * 100
  ) %>%
  select(Year, WOZ_index, Inkomen_index) %>%
  pivot_longer(
    cols = c(WOZ_index, Inkomen_index),
    names_to = "Variable",
    values_to = "Index"
  ) %>%
  mutate(
    Variable = recode(
      Variable,
      WOZ_index = "Average WOZ value",
      Inkomen_index = "Employee income"
    )
  )

ggplot(landelijk_index, aes(x = Year, y = Index, color = Variable)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Development of House Values and Employee Income in the Netherlands",
    subtitle = "Indexed values, 2015 = 100",
    x = "Year",
    y = "Index value",
    color = "Variable"
  ) +
  theme_minimal()+
  scale_x_continuous(breaks = seq(2015, 2024, by =1))


```

## 3.4 Visualize spatial variation

Here you provide a description of why the plot above is relevant to your specific social problem.

```{r Spacial Plot, echo=FALSE, message=FALSE, warning=FALSE}
provincie_map <- cbs_get_sf("provincie", 2023, verbose = TRUE)

provincie_samengevoegd <- provincie_samengevoegd %>%
  mutate(
    statcode = trimws(RegioS),
    AffordabilityRatio = GemiddeldeWOZ / InkomenWerknemer
  )

provincie_map <- provincie_map %>%
  mutate(
    statcode = trimws(statcode)
  )

provincie_map_data <- provincie_map %>%
  left_join(
    provincie_samengevoegd,
    by = "statcode"
  )

ggplot(provincie_map_data) +
  geom_sf(aes(fill = AffordabilityRatio), color = "white") +
  labs(
    title = "Housing Affordability by Province in 2023",
    subtitle = "Average WOZ value divided by employee income",
    fill = "WOZ-to-income ratio"
  ) +
  theme_minimal()
```
## 3.5 Visualize sub-population variation

What is the poverty rate by state?

Here you provide a description of why the plot above is relevant to your specific social problem.

```{r Subpopulation Boxplot, echo=FALSE, message=FALSE, warning=FALSE}
urban_affordability <- raw_dataset %>%
  mutate(
    Year = as.integer(substr(Perioden, 1, 4))
  ) %>%
  filter(
    Year == 2024,
    startsWith(RegioS, "GM")
  ) %>%
  select(
    Year,
    RegioS,
    RegioS_label,
    GemiddeldeWOZWaardeVanWoningen_98,
    BronInkomenAlsWerknemer_141,
    ZeerSterkStedelijk_52,
    SterkStedelijk_53,
    MatigStedelijk_54,
    WeinigStedelijk_55,
    NietStedelijk_56
  ) %>%
  rename(
    GemiddeldeWOZ = GemiddeldeWOZWaardeVanWoningen_98,
    InkomenWerknemer = BronInkomenAlsWerknemer_141
  ) %>%
  mutate(
    AffordabilityRatio = GemiddeldeWOZ / InkomenWerknemer
  ) %>%
  pivot_longer(
    cols = c(
      ZeerSterkStedelijk_52,
      SterkStedelijk_53,
      MatigStedelijk_54,
      WeinigStedelijk_55,
      NietStedelijk_56
    ),
    names_to = "UrbanizationType",
    values_to = "UrbanizationValue"
  ) %>%
  group_by(RegioS, RegioS_label, Year) %>%
  slice_max(UrbanizationValue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    UrbanizationType = recode(
      UrbanizationType,
      ZeerSterkStedelijk_52 = "Very strongly urban",
      SterkStedelijk_53 = "Strongly urban",
      MatigStedelijk_54 = "Moderately urban",
      WeinigStedelijk_55 = "Weakly urban",
      NietStedelijk_56 = "Non-urban"
    ),
    UrbanizationType = factor(
      UrbanizationType,
      levels = c(
        "Very strongly urban",
        "Strongly urban",
        "Moderately urban",
        "Weakly urban",
        "Non-urban"
      )
    )
  )

ggplot(urban_affordability, aes(x = UrbanizationType, y = AffordabilityRatio)) +
  geom_boxplot() +
  labs(
    title = "Housing Affordability by Urbanization Level",
    subtitle = "Municipalities grouped by dominant urbanization category, 2024",
    x = "Urbanization level",
    y = "WOZ-to-income ratio"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

## 3.6 Event analysis

Analyze the relationship between two variables.
Here you provide a description of why the plot above is relevant to your specific social problem.


```{r Affordability Index Plot, echo=FALSE}
landelijk_affordability_index <- landelijk %>%
  arrange(Year) %>%
  mutate(
    AffordabilityIndex = AffordabilityRatio / AffordabilityRatio[Year == 2015] * 100
  )

ggplot(landelijk_affordability_index, aes(x = Year, y = AffordabilityIndex)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed") +
  geom_vline(xintercept = 2024, linetype = "dashed", color = "red") +
  annotate(
    "text",
    x = 2023.85,
    y = max(landelijk_affordability_index$AffordabilityIndex, na.rm = TRUE)-27,
    label = "New Rent Act (2024)",
    hjust = 1,
    color = "red",
    size = 3.5
  ) +
  labs(
    title = "Housing Affordability Index in the Netherlands",
    subtitle = "Affordability ratio indexed to 2015 = 100",
    x = "Year",
    y = "Affordability index"
  ) +
  theme_minimal()+
  scale_x_continuous(breaks = seq(2015, 2024, by =1))
```

# Part 4 - Discussion

## 4.1 Discuss your findings

# Affordability ratio
# Growth gap

# Temporal variation
If we look at the temporal variation, which is constructed by indexing the average WOZ value and disposable income per employee using 2015 as a starting point, we see an alignment initially. However, after 2017 there starts to emerge a growth divergence, which spikes until 2023. Although a period of consolidation will follow, the growth gap is unlikely to close.

# Spatial variation
# Sub-population variation
# Event

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

1. De Nederlandsche Bank. (2026). Woningmarkt. https://www.dnb.nl/actuele-economische-vraagstukken/woningmarkt/

2. Centraal Bureau voor de Statistiek. (2024, 14 mei). Housing costs in the Netherlands higher than in most EU countries. https://www.cbs.nl/en-gb/news/2024/20/housing-costs-in-the-netherlands-higher-than-in-most-eu-countries


Use APA referencing throughout your document.