---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

# Set-up your environment

```{r Load Packages, message=FALSE, warning=FALSE}
library(cbsodataR)
library(dplyr)
library(tidyr)
library(sf)
library(knitr)
library(ggplot2)
```

# Title Page

Include your names
Melvin Hofman
Max Bergsma
Gijs Stevens
Hidde Voors
Imad el Hilali
Jules Bierings


Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Housing affordability has become an increasingly important social issue in the Netherlands over the past decade. House prices have risen substantially in many parts of the country, while income growth may not have kept pace. As a result, many households, particularly first-time buyers and younger generations, face greater difficulties entering the housing market. Rising housing costs can contribute to financial stress, wealth inequality, and reduced access to homeownership.

Understanding whether house prices have increased faster than incomes is therefore important for evaluating changes in housing affordability. If housing prices grow more rapidly than household incomes, purchasing a home becomes increasingly difficult for a large share of the population. This may have significant social and economic consequences, including increased inequality between homeowners and non-homeowners.

The main research question of this study is:

Have housing prices in the Netherlands increased faster than incomes over the past ten years?

To answer this question, the following sub-questions are examined:
1. How have house prices and average incomes developed between 2014 and 2024?
2. How has housing affordability changed during this period?
3. Are there significant differences in housing affordability between Dutch provinces?
4. Did the COVID-19 pandemic influence the relationship between house prices and income?

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r Download Data, message=FALSE, warning=FALSE}
# data_download <- cbs_get_data("70072NED")

# Add region label to the dataset
# raw_dataset <- data_download %>% cbs_add_label_columns()
# write.csv(raw_dataset, "data/raw_dataset.csv")

# Load saved dataset
raw_dataset = read.csv("data/raw_dataset.csv")
```


## 2.2 Provide a short summary of the dataset(s)

```{r Homevalue Dataset, message=FALSE, warning=FALSE}
huizenprijzen_provincie = raw_dataset %>%
  mutate(
    Year = as.integer(substr(Perioden, 1, 4))
  ) %>%
  filter(
    Year == 2023,
    startsWith(RegioS, "PV")
  ) %>%
  select(
    Year,
    RegioS,
    RegioS_label,
    GemiddeldeWOZWaardeVanWoningen_98
  )
write.csv(huizenprijzen_provincie, "data/huizenprijzen_provincie.csv")

kable(huizenprijzen_provincie)
```

```{r Income Dataset}
inkomen_provincie <- raw_dataset %>%
  mutate(
    Year = as.integer(substr(Perioden, 1, 4))
  ) %>%
  filter(
    Year == 2023,
    startsWith(RegioS, "PV")
  ) %>%
  select(
    Year,
    RegioS,
    RegioS_label,
    BronInkomenAlsWerknemer_141
  )
write.csv(inkomen_provincie, "data/inkomen_provincie.csv")

kable(inkomen_provincie)
```

```{r Province Datasets Merge}
provincie_samengevoegd = huizenprijzen_provincie %>%
  inner_join(
    inkomen_provincie,
    by = c("Year", "RegioS", "RegioS_label")
  ) %>%
  rename(
    GemiddeldeWOZ = GemiddeldeWOZWaardeVanWoningen_98,
    InkomenWerknemer = BronInkomenAlsWerknemer_141
  )  %>%
  mutate(statcode = RegioS)

kable(provincie_samengevoegd, digits = 2)
```

```{r Landelijke Dataset}
landelijk <- raw_dataset %>%
  mutate(
    Year = as.integer(substr(Perioden, 1, 4))
  ) %>%
  filter(
    Year >= 2015,
    Year <= 2024,
    startsWith(RegioS, "NL")
  ) %>%
  select(
    Year,
    RegioS_label,
    BronInkomenAlsWerknemer_141,
    GemiddeldeWOZWaardeVanWoningen_98
  ) %>%
  rename(
    GemiddeldeWOZ = GemiddeldeWOZWaardeVanWoningen_98,
    InkomenWerknemer = BronInkomenAlsWerknemer_141
  )

write.csv(landelijk, "data/landelijk.csv")

kable(landelijk)
```

```{r Summarize Data}
provincie_summary = provincie_samengevoegd %>%
  summarise(
    Mean_WOZ = mean(GemiddeldeWOZ),
    Min_WOZ = min(GemiddeldeWOZ),
    Max_WOZ = max(GemiddeldeWOZ),
    SD_WOZ = sd(GemiddeldeWOZ),
    
    Mean_Income = mean(InkomenWerknemer),
    Min_Income = min(InkomenWerknemer),
    Max_Income = max(InkomenWerknemer),
    SD_Income = sd(InkomenWerknemer),
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Statistic", "Variable"),
    names_sep = "_",
    values_to = "Value"
  ) %>%
  pivot_wider(
    names_from = Statistic,
    values_from = Value
  ) %>%
  mutate(
    Variable = recode(
      Variable,
      WOZ = "Average WOZ Value",
      Income = "Employee Income",
      Affordability = "WOZ-to-Income Ratio"
    )
  ) %>%
  select(Variable, Mean, Min, Max, SD)

kable(
  provincie_summary,
  caption = "Descriptive statistics for home value in provinces in the Netherlands, 2023",
  digits = 2
)
```
## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?


# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables


```{r New Variable 1}
provincie_samengevoegd <- provincie_samengevoegd %>%
  mutate(
    AffordabilityRatio = GemiddeldeWOZ / InkomenWerknemer
  )
provincie_samengevoegd

write.csv(provincie_samengevoegd, "data/provincie_samengevoegd.csv")
```
```{r New Variable 2}
landelijk = landelijk %>%
  mutate(
    AffordabilityRatio = GemiddeldeWOZ / InkomenWerknemer,
    WOZ_growth_yoy = (GemiddeldeWOZ / lag(GemiddeldeWOZ) - 1) * 100,
    Income_growth_yoy = (InkomenWerknemer / lag(InkomenWerknemer) - 1) * 100,
    GrowthGap_yoy = WOZ_growth_yoy - Income_growth_yoy
  )
kable(landelijk)
```



## 3.3 Visualize temporal variation

```{r Time Plot}
# Deze plot gemaakt met chat

landelijk_index <- landelijk %>%
  arrange(Year) %>%
  mutate(
    WOZ_index = GemiddeldeWOZ / GemiddeldeWOZ[Year == 2015] * 100,
    Inkomen_index = InkomenWerknemer / InkomenWerknemer[Year == 2015] * 100
  ) %>%
  select(Year, WOZ_index, Inkomen_index) %>%
  pivot_longer(
    cols = c(WOZ_index, Inkomen_index),
    names_to = "Variable",
    values_to = "Index"
  ) %>%
  mutate(
    Variable = recode(
      Variable,
      WOZ_index = "Average WOZ value",
      Inkomen_index = "Employee income"
    )
  )

ggplot(landelijk_index, aes(x = Year, y = Index, color = Variable)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Development of House Values and Employee Income in the Netherlands",
    subtitle = "Indexed values, 2015 = 100",
    x = "Year",
    y = "Index value",
    color = "Variable"
  ) +
  theme_minimal()+
  scale_x_continuous(breaks = seq(2015, 2024, by =1))


```

## 3.4 Visualize spatial variation

Here you provide a description of why the plot above is relevant to your specific social problem.

```{r Spacial Plot, message=FALSE, warning=FALSE}
provincie_map <- cbs_get_sf("provincie", 2023, verbose = TRUE)

provincie_samengevoegd <- provincie_samengevoegd %>%
  mutate(
    statcode = trimws(RegioS),
    AffordabilityRatio = GemiddeldeWOZ / InkomenWerknemer
  )

provincie_map <- provincie_map %>%
  mutate(
    statcode = trimws(statcode)
  )

provincie_map_data <- provincie_map %>%
  left_join(
    provincie_samengevoegd,
    by = "statcode"
  )

ggplot(provincie_map_data) +
  geom_sf(aes(fill = AffordabilityRatio), color = "white") +
  labs(
    title = "Housing Affordability by Province in 2023",
    subtitle = "Average WOZ value divided by employee income",
    fill = "WOZ-to-income ratio"
  ) +
  theme_minimal()
```
## 3.5 Visualize sub-population variation

What is the poverty rate by state?

Here you provide a description of why the plot above is relevant to your specific social problem.

```{r Subpopulation Boxplot, message=FALSE, warning=FALSE}
urban_affordability <- raw_dataset %>%
  mutate(
    Year = as.integer(substr(Perioden, 1, 4))
  ) %>%
  filter(
    Year == 2024,
    startsWith(RegioS, "GM")
  ) %>%
  select(
    Year,
    RegioS,
    RegioS_label,
    GemiddeldeWOZWaardeVanWoningen_98,
    BronInkomenAlsWerknemer_141,
    ZeerSterkStedelijk_52,
    SterkStedelijk_53,
    MatigStedelijk_54,
    WeinigStedelijk_55,
    NietStedelijk_56
  ) %>%
  rename(
    GemiddeldeWOZ = GemiddeldeWOZWaardeVanWoningen_98,
    InkomenWerknemer = BronInkomenAlsWerknemer_141
  ) %>%
  mutate(
    AffordabilityRatio = GemiddeldeWOZ / InkomenWerknemer
  ) %>%
  pivot_longer(
    cols = c(
      ZeerSterkStedelijk_52,
      SterkStedelijk_53,
      MatigStedelijk_54,
      WeinigStedelijk_55,
      NietStedelijk_56
    ),
    names_to = "UrbanizationType",
    values_to = "UrbanizationValue"
  ) %>%
  group_by(RegioS, RegioS_label, Year) %>%
  slice_max(UrbanizationValue, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    UrbanizationType = recode(
      UrbanizationType,
      ZeerSterkStedelijk_52 = "Very strongly urban",
      SterkStedelijk_53 = "Strongly urban",
      MatigStedelijk_54 = "Moderately urban",
      WeinigStedelijk_55 = "Weakly urban",
      NietStedelijk_56 = "Non-urban"
    ),
    UrbanizationType = factor(
      UrbanizationType,
      levels = c(
        "Very strongly urban",
        "Strongly urban",
        "Moderately urban",
        "Weakly urban",
        "Non-urban"
      )
    )
  )

ggplot(urban_affordability, aes(x = UrbanizationType, y = AffordabilityRatio)) +
  geom_boxplot() +
  labs(
    title = "Housing Affordability by Urbanization Level",
    subtitle = "Municipalities grouped by dominant urbanization category, 2024",
    x = "Urbanization level",
    y = "WOZ-to-income ratio"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

## 3.6 Event analysis

Analyze the relationship between two variables.
Here you provide a description of why the plot above is relevant to your specific social problem.


```{r Affordability Index Plot}
landelijk_affordability_index <- landelijk %>%
  arrange(Year) %>%
  mutate(
    AffordabilityIndex = AffordabilityRatio / AffordabilityRatio[Year == 2015] * 100
  )

ggplot(landelijk_affordability_index, aes(x = Year, y = AffordabilityIndex)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed") +
  labs(
    title = "Housing Affordability Index in the Netherlands",
    subtitle = "Affordability ratio indexed to 2015 = 100",
    x = "Year",
    y = "Affordability index"
  ) +
  theme_minimal()+
  scale_x_continuous(breaks = seq(2015, 2024, by =1))
```

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.